Getting and installing R
---------------

### To start:

1. Download and install **R** from [https://cran.r-project.org/](https://cran.r-project.org/)
2. Download and install **RStudio** from [https://www.rstudio.com/](https://www.rstudio.com/)
3. Open RStudio
4. Print to the editor:

```{r setwdex, eval=FALSE}
# Sets working directory, into which things are saved by default and from which stuff is imported
setwd("/Users/ulomaivali/Dropbox/") # mac
setwd("C:/Users/ulomaivali/Dropbox/") # windows
setwd("~/Dropbox/R course/") # linux, ~/ references to user home directory
```

```{r getwd}
# Prints the current working directory
getwd()
```

### Getting help

You can get help from R Documentation at the program's command prompt by preceding R command with ?.

```{r help}
?getwd()
```

### Random tips

* `Cmd + Enter` (`Ctrl + Enter` on a PC) in RStudio: sends the current line (or current selection) from the editor to the console and runs it. 
* R is case-sensitive. 
* Enter commands one at a time at the console or run a set of commands from the editor. 
* Object names cannot begin with a number. 
* No spaces in object names! (use . or _ or -).
