---

---

Selecting by index
------------------

+ index is a quick address used to select/filter elements of vector/matrix/df/list
+ first rows, then columns: [1, 2] means "select 1st row, 2nd column"
+ indexing begins at 1 (not 0)
+ index is in square brackets (list elements are selected by double square brackets). '[' and '[[' are functions!

```{r}
'['(c(1,2,3), 1)
```

+ you can select values, rows or columns by index
+ empty means "all" eg. `a[,1]` select all rows and 1st column from a
+ `-` (minus) before index means "not this element"
+ ! means "not"
+ != means "not equal"
+ == means "equal"

```{r, eval = FALSE}
df[ , 2] # selects 2nd column
df[2, ] # selects 2nd row
df[1:10, 2:3]
```
```{r}
df <- data.frame(M = c(2,3,6,1,34), N = c(34,1,8,9,3))
df
df[df$M == 34, ] # selects rows from A that have value == 34
```
```{r}
df[1:2, "N"] # selects rows 1 through 2 from column "A"
```

```{r}
rownames(df) <- letters[1:5]
df[rownames(df)=="c", ] # selects row named "c"
```

```{r}
df[-(2:4), ] # drops rows 2 to 4 (incl)
```

```{r}
df[,-2] # drops col 2, outputs vector! 
```


df[ !(X==”a”), ] selects all rows that do not contain “a” in column named X

df[AA==T, ] selects all rows where AA is TRUE

df[! duplicated(X), ] selects 1st value of duplicated values in column named X

df[col1 > median(col1) & col2 < 5.2, ] selects rows where col1 value is > median & col2 value < 5.2

df[col1 > 8 | col2 < 5.2, ] selects rows where col1 value is > 8 or col2 < 5.2

sum(df$a[df$sex==”male”]) – sums column “a” at rows where column sex is “male” 