---

---

Selecting by index
------------------
TODO: lihtsamaks!!!

+ index is a quick address used to select/filter elements of vector/matrix/df/list
+ first rows, then columns: [1, 2] means "select 1st row, 2nd column"
+ indexing begins at 1 (not 0)
+ index is in square brackets (list elements are selected by double square brackets). '[' and '[[' are functions!

```{r}
'['(c(1,2,3), 1) == c(1,2,3)[1] # as you can see these are both identical
```

+ you can select values, rows or columns by index
+ empty means "all" eg. `a[,1]` select all rows and 1st column from a
+ `-` (minus) before index means "not this element"
+ ! means "not"
+ != means "not equal"
+ == means "equal"

```{r, eval = FALSE}
df[,2] # selects 2nd column
df[2,] # selects 2nd row
df[1:10, 2:3]
```
```{r}
df <- data.frame(M = c(2, 3, 6, 3, 34), N = c(34, 3, 8, 3, 3), L = c(TRUE, FALSE, TRUE, FALSE, TRUE))
df
df[df$M == 34,] # selects rows from A that have value == 34
```
```{r}
df[1:2, "N"] # selects rows 1 through 2 from column "A"
```

```{r}
rownames(df) <- letters[1:5]
df[rownames(df) == "c",] # selects row named "c"
```

```{r}
df[-(2:4),] # drops rows 2 to 4 (incl)
```

```{r}
df[, -2] # drops col 2, outputs vector! 
```

```{r}
df[df$M==6,] # selects all rows that do not contain 6 in column named M
```


```{r}
df[df$L==T,] # selects all rows where L is TRUE (T)
```

```{r}
df[!duplicated(df),] # removes second one of the duplicated rows from df, we have to use ! to negate logical evaluation
```

```{r}
df[df$M > median(df$N) & df$M < 25,] # selects rows where df$M value is > median df$N AND df$M value < 25
```


```{r}
df[df$M > median(df$N) | df$M == 34,] # selects rows where df$M value is > median df$N OR df$M value == 34
```

```{r}
sum(df$M[df$L==T]) # sums column df$M at rows where column 'L' is TRUE (T)
```

