---

---

# Selecting by index

+ Index gives the address that specifies the elements of vector/matrix/list or data.frame, which are then automatically selected.
+ Indexing begins at 1 (not 0) in R
+ Indexing operators in R are square brackets -- '[', `[[` and dollar sign `$`  
+ `[` allows selecting more than one element, whereas `[[` and `$` select only one element.
+ Empty index [,] means "select all" -- `a[,1]` means "select all rows and 1st column from `a`".

<!-- ```{r} -->
<!-- '['(c(1,2,3), 1) == c(1,2,3)[1] # as you can see these are both identical -->
<!-- ``` -->

<!-- + you can select values, rows or columns by index -->

<!-- + `-` (minus) before index means "not this element" -->
<!-- + ! means "not" -->
<!-- + != means "not equal" -->
<!-- + == means "equal" -->

During initial data exploration it is often necessary to have a look how the head of your table looks like, for this you can use convenience methods `head` and `tail` which are returning first and last elements of a object, respectively:
```{r}
head(mtcars) # Prints first 6 elements (rows) as default
tail(mtcars, n = 3) # Prints last 3 elements (rows)
```

Tip: you can use `tail` to return the very last element of a object with unknown length.

```{r}
tail(LETTERS, n = 1)
```

## Indexing matrices
Matrices are vectors with a dimension attribute and so all the vector forms of indexing can be used with a single index. 

```{r}
m <- matrix(1:12, nrow = 3, dimnames = list(c("a","b","d"), LETTERS[1:4]))
m
```

```{r}
m[1,]                 # the first row of matrix m
```

```{r}
m[1, , drop = FALSE]  # is a 1-row matrix
```

```{r}
m[, c(TRUE, FALSE, FALSE, TRUE)] # logical indexing
```

```{r}
m[cbind(c(1,2,1), 3:1)] # matrix numeric index
```

```{r}
ci <- cbind(c("a", "b", "a"), c("A", "C", "B"))
m[ci]                 # matrix character index
```

```{r}
m[,-1]           # delete the first column of m
```


## Indexing lists
Indexing by `[` is similar to atomic vectors and selects a list of the specified element(s).
Both `[[` and `$` select a single element of the list (e.g. a single vector or data frame).

```{r}
mylist # here we go back to our mylist object
```

```{r}
mylist[[1]]               # the first element of list mylist
```

```{r}
mylist[c(1, 3)]           # a list containing elements 1 and 3 of mylist
```

```{r}
mylist$ages               # the element of mylist named ages
```

<!-- ### Recursive indexing into lists -->
<!-- ```{r} -->
<!-- z <- list(a = list(x = 9, y = "chello"), b = 1:5) # we create list named 'a' inside list 'z' -->
<!-- z -->
<!-- ``` -->

<!-- ```{r} -->
<!-- unlist(z) # look what happens if we simplify this list into vector... -->
<!-- ``` -->

<!-- ```{r} -->
<!-- z$a$x # what's a$x in list z -->
<!-- ``` -->


<!-- ```{r} -->
<!-- z[[c(1, 2)]] # we get second element of 1st element of list z -->
<!-- ``` -->
<!-- ```{r} -->
<!-- z[[c("a", "x")]] <- "piano" # here we replace z$a$x with "piano" -->
<!-- z -->
<!-- ``` -->


## Indexing data.frames
<!-- Data frames can be indexed in several modes. When [ and [[ are used with a single vector index (x[i] or x[[i]]), they index the data frame as if it were a list. -->

We use R `mtcars` dataset to illustrate indexing of a data.frame:
```{r}
class(mtcars)
dim(mtcars) # what's the size of the data.frame
mtc <- mtcars[sample(1:nrow(mtcars), 6), ] # select a manageable subset
mtc
```

Here we select columns: 
```{r}
mtc[,2] # selects 2nd column and returns vector
mtc[3] # selects 3nd column and returns data.frame
mtc[, "hp"] # selects column named "hp"
mtc$cyl # selects column named "cyl"
```


```{r}
df <- data.frame(M = c(2, 3, 6, 3, 34), N = c(34, 3, 8, 3, 3), L = c(TRUE, FALSE, TRUE, FALSE, TRUE))
df
df[df$M == 34,] # selects rows from A that have value == 34
```
```{r}
df[1:2, "N"] # selects rows 1 through 2 from column "A"
```

```{r}
rownames(df) <- letters[1:5]
df[rownames(df) == "c",] # selects row named "c"
```

```{r}
df[-(2:4),] # drops rows 2 to 4 (incl)
```

```{r}
df[, -2] # drops col 2, outputs vector! 
```

```{r}
df[df$M==6,] # selects all rows that do not contain 6 in column named M
```


```{r}
df[df$L==T,] # selects all rows where L is TRUE (T)
```

```{r}
df[!duplicated(df),] # removes second one of the duplicated rows from df, we have to use ! to negate logical evaluation
```

```{r}
df[df$M > median(df$N) & df$M < 25,] # selects rows where df$M value is > median df$N AND df$M value < 25
```


```{r}
df[df$M > median(df$N) | df$M == 34,] # selects rows where df$M value is > median df$N OR df$M value == 34
```

```{r}
sum(df$M[df$L==T]) # sums column df$M at rows where column 'L' is TRUE (T)
```

A vector can be extracted by `$` and worked on:

```{r}
Mean.height <- mean(my_tab$height)
Mean.height # Prints the answer
```

New vectors can be bound into a data.frame:

```{r}
my_tab$weight <- c(87, 96, 69) # Now there are 3 columns in my_tab
my_tab
```


```{r}
my_tab$experiment <- factor("A") # the 4th col contains a factor with a single level "A"
levels(my_tab$experiment) # prints the unique levels in a factor vector
```
