---

---

# ggplot2
**ggplot2** is an R package for producing statistical graphics based on the grammar of graphics (hence the gg!).

+ **ggplot2 works iteratively** -- you start with a layer showing the raw data and then add layers of geoms, annotations, and statistical summaries.

To compose plots, you have to supply minimally:

+ **Data** that you want to visualise and **aes**thetic mappings -- what's on x-axis, what's on y-axis, and how to you want to group and color your data.
+ **Layers** made up of **geom**etric elements: points, lines, boxes, etc. 

You can further adjust your plot:

+ by adding **statistical summaries** of your raw data.
+ using **scale**s to redraw a legend or axes.
+ using **facet**ing to break up the data into subsets for display.
+ using **theme**s which control plot features like the font size 
  and background colour.

ggplot2 is different from base graphics:

+ Plots created by base graphics appear only on the screen and you cannot assign plot to an object for later use. Everything is created in place.
+ You can only draw on top of the plot, you cannot modify or delete existing content.

That was theory, you can read more from [ggplot2-book](https://github.com/hadley/ggplot2-book), this is where rubber meets the road:
```{r}
library(ggplot2) # load ggplot2 library
library(dplyr) # dplyr is necessary for piping operator and for some data munging
library(tidyr)
```

We use ggplot2 builtin dataset `mpg` with fuel economy data from 1999 and 2008 for 38 models of car:
```{r}
mpg
```
Key variables in `mpg` dataset are `displ` -- engine displacement (L), `cyl` -- number of cylinders, `cty` -- city miles per gallon, `hwy` -- highway miles per gallon.

Simple scatterplot to explore relationship between fuel consumption in city traffick (cty) and engine size (displ) is created like this:
```{r}s
ggplot(mpg, aes(displ, cty)) + geom_point()
```
**Notice that**: in ggplot first two unnamed arguments to `aes()` are `x` and `y`, so we can easily skip `x =` and `y =` to reduce typing.

To illustrate layerwise plot creation in ggplot:
```{r}
p <- ggplot(data = mpg) # first add data
p <- p + aes(x = cty, y = hwy) # add aesthetics
p <- p + geom_point() # add geom, specifying plot type 
p <- p + stat_smooth(method = "lm") # add statistical summary
p # display plot
```
The **point geom is used to create scatterplots**.

Of course, same result is easily achieved by this oneliner:
```{r, eval=FALSE}
ggplot(mpg, aes(cty, hwy)) + geom_point() + stat_smooth(method = "lm")
```

## Histograms
Differently from base R graphics, where histogram is drawn by a special command, ggplot specifies the type of plot by geoms, like `geom_histogram()`:
```{r}
df <- tibble(x = rnorm(10000)) # For demo we create tibble with 10000 random numbers from normal distribution
ggplot(df, aes(x)) + geom_histogram() # to plot histogram we specify geom_histogram 
```

ggplot says: `stat_bin()` using `bins = 30`. Pick better value with `binwidth`, also help file notes that 'the default is to use bins bins that cover the range of the data. You should always override this value, exploring multiple widths to find the best to illustrate the stories in your data'. OK, Messy Goes to Okido, lets try:
```{r}
ggplot(df, aes(x)) + geom_histogram(binwidth = 0.1) # for better granularity, we specify binwith = 0.1
```

Alternatively, you can specify the number of bins (and perhaps send some hidden messages along with your visualisation):
```{r}
ggplot(df, aes(x)) + geom_histogram(bins = 5) # if we want to split data into eg. 5 bins
```

For some more practice, lets recreate Tartu Marathon 2013 histogram:
```{r}
library(lubridate)
load("data/Tartu_Maraton_2013.RData")
```

```{r}
tm <- tm_2013 %>% mutate(Aeg = unclass(as.duration(hms(Aeg)))/3600) %>% filter(complete.cases(.))
tm %>% ggplot(aes(Aeg)) + geom_histogram(binwidth = 1/12) # we want binwidth ~5 minutes (1/12 hour)
```

These were histograms with counts on y-axis, but we may want to plot (kernel) density instead, for this we use `geom_density()`:
```{r}
ggplot(df, aes(x)) + geom_density()
```

Histogram with densities on y-axis:
```{r}
ggplot(df, aes(x, ..density..)) + geom_histogram()
```

Kernel density plot and histogram together: 
```{r}
ggplot(df, aes(x, ..density..)) + 
  geom_histogram(binwidth = 0.1) +
  geom_density()
```

```{r}
p <- ggplot(df, aes(x)) + 
  geom_histogram(aes(y = ..density..), # note that we specify density here, 
                 binwidth = 0.1, 
                 fill = "white", color = "black") +
  geom_density()
p # we disrected plot to object
```

We can add vertical line, denoting mean, to our histogram like this:
```{r}
p + geom_vline(aes(xintercept = mean(x)), color = "red", linetype = "dashed", size = 1) 
```

**To compare distributions** we can overlay histograms and density plots. To demonstrate this, let's create dummy dataset of two normal distributions:
```{r}
df2 <- tibble(pop = rep(c("N", "S"), each = 1000), 
              value = c(rnorm(1000, 20, 3), rnorm(1000, 25, 3)))
df2
```

When overlaying histograms in ggplot, default action is to 'stack' them one to another, making comparison difficult. We have to specify `position = "identity"` to place them directly onto x-axis (each bar starts from y = 0). Also we specify `alpha =` for some transparency:
```{r}
p1 <- df2 %>% ggplot(aes(value, fill = pop)) 
p1 + geom_histogram(position = "identity", alpha = 0.7, binwidth = 0.3)
```

Alternatively, it's possible to place comparable bars next to each other by specifying `position = "dodge"`:
```{r}
p1 + geom_histogram(position = "dodge", binwidth = 1)
```

Overlaid kernel density plots:
```{r}
p2 <- p1 + geom_density(alpha = 0.7)
p2
```

To indicate means for these two distributions, we have to calculate means into a new data frame: 
```{r}
df2_stats <- df2 %>% group_by(pop) %>% summarise(pop_mean = mean(value))
p2 + geom_vline(data = df2_stats, aes(xintercept = pop_mean, color = pop), linetype = "dashed", size = 1)
```

##  Boxplots

**Boxplots** are created using `geom_boxplot`:
```{r}
df2 %>% ggplot(aes(pop, value)) + geom_boxplot()
```

```{r}
p3 <- df2 %>% ggplot(aes(pop, value, fill = pop)) + geom_boxplot()
p3
```

Often, it's a good idea to overlay original values to plot. Specially, if there is only few values. Let's add original data points to previous boxplot:
```{r}
p3 + geom_point()
```
Adding geom_points was not very helpful in this case.
s
We can try to jitter these points, where argument `width=` specifies jitter width (default full boxplot width):
```{r}
p3 + geom_jitter(width = 0.5)
```

As shown previously, we can also add summary data (mean) to this plot (large red dots):
```{r}
p3 + geom_point(data = df2_stats, aes(y = pop_mean), shape = 16, color = "red", size = 3)
```


## Point shape codes
It seems impossible to remember available point shape codes in R. The quickest way out of this is to know how to generate an example plot of the shapes quickly. [This is how to do this in ggplot2.](http://www.win-vector.com/blog/2012/04/how-to-remember-point-shape-codes-in-r/)
```{r}
ggplot(data=data.frame(x=c(0:25))) + geom_point(size=8, aes(x=x,y=x,shape=x)) +
  facet_wrap(~ x, scales='free') + xlab('') + ylab('') +
  scale_shape_identity() +
  theme(axis.text.x=element_blank(), axis.text.y=element_blank(),
        axis.ticks=element_blank(), legend.position='none',
        panel.background=element_blank(),
        panel.grid.major=element_blank(),
        panel.border=element_blank())
```

## Plotting means and error bars
Summaries and e.g. errorbars can also be calculated on the fly. To illustrate this, we take first smaller sample from our dataset in order to have larger errorbars. Here we add bootstrapped 99% confidence intervals of the mean:
```{r}
p4 <- df2 %>% sample_n(50) %>% # to have larger error
  ggplot(aes(pop, value, fill = pop)) + geom_jitter(color = "gray", width = 1/3) 
p4 + stat_summary(fun.data = "mean_cl_boot", fun.args = list(conf.int = .99), geom = "pointrange") # poinrange is default geom 
```

Here we add errorbars, denoting standard error of the mean:
```{r}
p4 <- p4 + stat_summary(fun.y = mean, geom = "point", shape = 95, color = "red", size = 15) + # mean
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25) # standard deviation
p4
```

## Start y or x axis at zero
In many cases proper visual representation of data requires that e.g. y axis starts from zero. By default, ggplot zooms only into region where data is located and 0 is dropped from axes. To change this behaviour, ggplot has command `expand_limits()` specifying the value(-s) that should be included in each scale:
```{r}
p4 + expand_limits(y = 0) # upper and lower axis limits can be specified by supplying vector of two values
```

`scale_x_continuous()` and `scale_y_continuous()` have argument limits, which can be used to set axis limits, 
```{r}
p4 + scale_y_continuous(limits = c(15, 20))
```


## Linegraphs
Now we create line graphs with errorbars, we use `mpg` data. We calculate mean and standard error of the mean for highway fuel consumption (hwy) per year and per number of cylinders:
```{r}
mpg_sum <- mpg %>% 
  filter(cyl %in% c(4,6,8)) %>%  # to include only comparable values
  group_by(year, cyl) %>% # to compare two years
  summarise(mean_hwy = mean(hwy), # to calculate mean
            N = n(), # N to calculate standard error
            se_hwy = sd(cty)/sqrt(N)) # standard error
mpg_sum
```

`geom_line()` is used to create linegraphs in ggplot. We want two distinct colors for year variable, therefore we coerce year to factor:
```{r}
p5 <- mpg_sum %>% ggplot(aes(cyl, mean_hwy, color = factor(year))) 
p5 + geom_point() + geom_line()
```

Errorbars can be added by using `geom_errorbar`, which needs arguments `ymin` and `ymax`:
```{r}
p5 + geom_point() + geom_line() +
  geom_errorbar(aes(ymin = mean_hwy + se_hwy, ymax = mean_hwy - se_hwy), width = 0.25)
```

When the errorbars overlap, use `position_dodge` to move them horizontally.
```{r}
pd <- position_dodge(0.7) # move them 
```

```{r}
p5 + geom_point(position = pd) + 
  geom_line(position = pd) +
  geom_errorbar(aes(ymin = mean_hwy + se_hwy, ymax = mean_hwy - se_hwy), 
                   width = 0.25, 
                   position = pd)
```

## Barplots
Barplot are created similarly to linegraphs, but with `geom_bar` and default stat `counts` needs to changed. Unstacked bars need also `position = position_dodge()` specification:
```{r}
pd <- position_dodge(0.9) 
p6 <- mpg_sum %>% ggplot(aes(factor(cyl), mean_hwy, fill = factor(year)))
p7 <- p6 + geom_bar(stat = "identity", position = pd)
p7
```
Note that bargraphs start at 0.


```{r}
p7 + geom_errorbar(aes(ymin = mean_hwy - se_hwy, ymax = mean_hwy + se_hwy),
                   position = pd, width = 0.25)
```


