---

---

# ggplot2
**ggplot2** is an R package for producing statistical graphics based on the grammar of graphics (hence the gg!).

Main features of ggplot2:

+ **ggplot2 works iteratively** -- you start with a layer showing the raw data and then add layers of annotations and statistical summaries.
+ **default** plots include **legend** and are styled using **high data-ink ratio theme**.

To compose a plots, you have to supply minimally:

+ **Data** that you want to visualise and **aes**thetic mappings -- what's on x-axis, what's on y-axis, and how to you want to group and color your data.
+ **Layers** made up of **geom**etric elements: points, lines, boxes, etc. 

You can further adjust your plot:

+ by adding **statistical summaries** of your raw data.
+ using **scale**s to redraw a legend or axes.
+ using **facet**ing to break up the data into subsets for display.
+ using **theme**s which control plot features like the font size 
  and background colour.

ggplot2 is different from base graphics:

+ Plots created by base graphics appear only on the screen -- you cannot assign plot to an object for later use. Everything is created in place.
+ You can only draw on top of the plot, you cannot modify or delete existing content.

```{r}
library(ggplot2)
library(dplyr)
```

ggplot2 has builtin dataset `mpg` with fuel economy data from 1999 and 2008 for 38 popular models of car:
```{r}
mpg
```

Key variables: displ -- engine displacement (L), cyl -- number of cylinders, cty -- city miles per gallon, hwy -- highway miles per gallon:
```{r}
summary(mpg)
```

Illustration of iterativity of ggplot creation:
```{r}
p <- ggplot(data = mpg) # first add data
p <- p + aes(x = cty, y = hwy) # add aesthetics
p <- p + geom_point() # add geom
p <- p + stat_smooth(method = "lm") # add statistical summary
p # display plot
```

More commonly plots are created like this:
```{r}
mpg %>% ggplot(aes(x = factor(cyl), y = cty)) + geom_point()
```
`aes()` is inside ggplot call. Note that dataset was piped into ggplot and `geom_point()` layer was added (`+`) to the ggplot.
