---
---

# Merging tables

## Combine by rows and columns
Base R `rbind` and `cbind` **take in a sequence of vector, matrix or data-frame** and combine by rows or columns, respectively. Meaning that if you want to combine only data frames you have also option to use Hadleyverse's `dplyr::bind_rows` or `dplyr::bind_cols`. Otherwise, if playing around with vectors and/or matrices too, you have to stick with base R functions.  

Don't try to understand this excerpt from help: for `rbind` column names are taken from the first argument with appropriate names: colnames for a matrix, or names for a vector of length the number of columns of the result.

To illustrate combining two data.frames by rowwise we first subset `mtcars` dataset into two
mini dataframes:
```{r}
dim(mtcars)
df1 <- mtcars[1:2, c(1,4)]
df1
```

```{r}
df2 <- mtcars[11:13, c(1,4)]
df2
```

Combine by rows using base R `rbind`, **columns need to match**:
```{r}
rbind(df1, df2)
```

Hadleyverse's verb to perform this task is `bind_rows`, however note that `bind_rows` (`bind_cols`) **needs data frames** (NO vectors or matrices!): 
```{r}
library(dplyr)
bind_rows(df1, df2) 
```

Neat feature of the `bind_rows` function is that when you supply a column name with the `.id=` argument, a new column is created to link each row to its original data frame:
```{r}
bind_rows(df1 = df1, df2 = df2, .id = "id")
```

**Columns don't need to match** when row-binding using `bind_rows`, missing values are filled with `NA`-s:
```{r}
bind_rows(df1 = df1, 
          df2 = mtcars[11:13, c(1,4,6)], 
          df3 = mtcars[4:6, c(1,6)],
          .id = "id")
```

If you have **data frames or tibbles in a list**, then you can bind them rowwise into one data frame. 

Here we first create list with two dataframes:
```{r}
df_list <- list(df1, df2)
df_list
```

`do.call(rbind, my_list_of_dataframes)` is base R solution to this task, but you need data frames with same column names:
```{r}
do.call(rbind, df_list) # bind dataframes in list by rows
```

`bind_rows` argument can also be a **list of data frames**, and again, columns are matched by name, and any values that don't match will be filled with `NA`-s:
```{r}
df_list2 <- list(df1 = df1, 
          df2 = mtcars[11:13, c(1,4,6)], 
          df3 = mtcars[4:6, c(1,6)])
bind_rows(df_list2, .id = "id")
```

Combining two vectors of same length columnwise results in matrix (similar to rbind):
```{r}
cbind(letters[1:5], 1:5)
```

When combining vectors with different lengths, the shorter one is recycled:
```{r}
m <- cbind(1, 1:7) # the '1' (= shorter vector) is recycled
m
rbind(1:6, 7:8) # same applies to rbind
```

Columns can be inserted by specifing order of colums in the index:
```{r}
cbind(m, 8:14)[, c(1, 3, 2)] # insert a column
```

When the arguments consist of a mix of matrices and vectors the number of rows of the result is determined by the number of rows of the matrix arguments:
```{r}
cbind(1:7, matrix(11:16, nrow = 3)) # vector is subset -> warning
```

When column-binding, rows are matched by position, not value so all data frames must have the same number of rows:
```{r}
bind_cols(df1, mtcars[1:2,5:6])
```

Summary:

+ `rbind`, `cbind` take in vectors, matrices and data frames.
+ `bind_rows`, `bind_cols` -- **data frames only**.
+ with `rbind` column names must match as well numbers of columns.
+ `bind_rows` handles non-matching cols and rows with ease.
+ in case of combining vector with array, `cbind` handles non-matching vector length with row number and recycles or subsets vector if necessary.
+ `bind_cols` need same number of rows in data frames.
+ whatever you combine, onus is on you to ensure that proper rows and columns get combined, when they get combined.

## Joins
During data analysis you find out quite often that your data is lying around in multiple tables and you need to combine these sets.
In `dplyr`, [there are three families of verbs](https://cran.r-project.org/web/packages/dplyr/vignettes/two-table.html) that work with two tables at a time:

+ Mutating joins, which add new variables to one table from matching rows in another.
+ Filtering joins, which filter observations from one table based on whether or not they match an observation in the other table.
+ Set operations, which combine the observations in the data sets as if they were set elements.

To illustrate joining, let's consider [following data](http://www.ester.ee/record=b2697405*est). 
Here we have day of travel, city, and population.
```{r}
library(dplyr)
trip <- data_frame(day = c(1:8), city = c("Teheran", "Qom", "Esfahan", "Persepolis", "Shiraz", "Bandar", "Karaj", "Tabriz"), pop = c(8846782,1374036,1945765,NA,1460665,212,1967005,1549453))
head(trip)
```

Next table we have hotels with addresses for some of the cities.
```{r}
hotels <- data_frame(city = c("Teheran", "Qom", "Esfahan", "Persepolis", "Shiraz", "Bandar"), hotel = c("Firouzeh Hotel", "Aria Hotel", "Amir Kabir Hotel", "Sita Aukh Guest House", "Anvari Hotel", "Bolivar Inn"), address = c("Dowlat Abad Alley", "Astane Square", "Chahar Bagh Abbasi Street", "Tume Junn Mohammad Street", "Anvari Street", "Abuzar Street"))
head(hotels)
```

### Mutating joins
We use `dplyr` `left_join` and `inner_join` functions to combine these tables by explicitly specifying column/variable 'by = "city"' (e.g. in order to guarantee consistent behaviour when number of columns in our data frame can vary in our script). If keep default 'by = NULL', then two datasets are combined by all matching columns.

```{r include=FALSE}
x <- as.character(bquote(trip))
y <- as.character(bquote(hotels))
```

`left_join` returns all rows from `r paste(x)` where there are matching values in `r paste(x)`, and all columns from `r paste(x)` and `r paste(y)`.
```{r}
left_join(trip, hotels, by = "city") 
```
`left_join` is the most commonly used join because it ensures that you don't lose observations from your primary table on the left side.

`inner_join` returns all rows from `r paste(x)`, and all columns from `r paste(x)` and `r paste(y)`. Rows in `r paste(x)` with no match in `r paste(y)` will have NA values in the new columns.
```{r}
inner_join(trip, hotels, by = "city") 
```

<!-- Lets consider immune assay data on a 8-well slide, we have quantitated whole plate, but we performed our assay only in wells 3 to 6: -->
<!-- ```{r} -->
<!-- well <- c(outer(LETTERS[1:4], 1:2, paste0)) # microplate wells rows A to H, and cols 1 to 12  -->
<!-- slide <- data_frame(well, value = runif(8)) # we have 'read in' some values for each well -->
<!-- slide -->
<!-- ``` -->

<!-- ```{r} -->
<!-- metadata <- data_frame(well = well[3:6],  -->
<!--                        Ab = rep(c("catb","cdh1"), 2)) -->
<!-- metadata -->
<!-- ``` -->

<!-- `inner_join` only includes observations that match in both x and y. -->
<!-- ```{r} -->
<!-- inner_join(slide, metadata)  -->
<!-- ``` -->

<!-- ```{r} -->
<!-- left_join(slide, metadata) -->
<!-- ``` -->

```{r}
df1 <- data_frame(x = c(1, 2), y = c(T, F))
df1
```

```{r}
df2 <- data_frame(x = c(1, 3), a = 10, b = "a")
df2
```

`left_join(x, y)` includes all observations from x and observations from y that have match in x:
```{r}
left_join(df1, df2)
```

`inner_join(x, y)` only includes observations that match in both x and y:
```{r}
inner_join(df1, df2)
```

Return all rows from `r paste(x)` where there are not matching values in `r paste(y)`, keeping just columns from `r paste(x)`.
```{r}
anti_join(trip[1:3,], hotels[3:4,], by = "city") # returns cols fron trip where there are no matching values in hotels 
```

Return all rows and all columns from both `r paste(x)` and `r paste(y)`. Where there are not matching values, returns NA for the one missing.
```{r}
full_join(trip[1:3,], hotels[3:6,], by = "city") 
```
