---

---

# Merging tables

## Combine by rows
For rbind column names are taken from the first argument with appropriate names: colnames for a matrix, or names for a vector of length the number of columns of the result.

To illustrate combining two data.frames by rowwise we first split `mtcars` dataset into two
```{r}
dim(mtcars)
head(mtcars)
df1 <- mtcars[1:4, ]
df2 <- mtcars[11:14, ]
```

```{r}
# a <- rbind(df1, df2)
b <- dplyr::bind_rows("group 1" = df1, "group 2" = df2, .id = "groups") # When you supply a column name with the `.id` argument, a new column is created to link each row to its original data frame. Columns don't need to match when row-binding
b
```

## Joins
To illustrate merging, let's consider [following itinerary](http://www.ester.ee/record=b2697405*est). 
Here we have day of travel and city.
```{r}
itinerary <- data.frame(day = c(1:6), city = c("Teheran", "Qom", "Esfahan", "Persepolis", "Shiraz", "Bandar"))
head(itinerary)
```

Next table we have hotels with addresses for each city.
```{r}
lodging <- data.frame(city = c("Teheran", "Qom", "Esfahan", "Persepolis", "Shiraz", "Bandar"), hotel = c("Firouzeh Hotel", "Aria Hotel", "Amir Kabir Hotel", "Sita Aukh Guest House", "Anvari Hotel", "Bolivar Inn"), address = c("Dowlat Abad Alley", "Astane Square", "Chahar Bagh Abbasi Street", "Tume Junn Mohammad Street", "Anvari Street", "Abuzar Street"))
head(lodging)
```

<!-- #### base::merge -->
<!-- We use base R function `merge` to join these tables by explicitly specifying column/variable 'by = "city"' (e.g. in order to guarantee consistent behaviour when number of columns in our data frame can vary in our script). -->
<!-- By default, `merge` joins data frames on the columns with names they both have. -->

<!-- ```{r} -->
<!-- merge(itinerary, lodging, by = "city", sort = FALSE) # don't sort on 'by' column -->
<!-- ``` -->

<!-- If we choose only rows 1 through 3 from table `lodging`? -->
<!-- ```{r} -->
<!-- merge(itinerary, lodging[1:3,], by = "city", sort = FALSE)  -->
<!-- ``` -->

<!-- If we choose only rows 2 through 5 from table `itinerary`? -->
<!-- ```{r} -->
<!-- merge(itinerary[2:5,], lodging, by = "city", sort = FALSE)  -->
<!-- ``` -->

<!-- In our simple example we always have only matching rows in our results. -->

<!-- #### dplyr::inner_join and friends -->
We use `dplyr` `_join` functions to join these tables by explicitly specifying column/variable 'by = "city"' (e.g. in order to guarantee consistent behaviour when number of columns in our data frame can vary in our script).

```{r include=FALSE}
x <- as.character(bquote(itinerary))
y <- as.character(bquote(lodging))
```


Returns all rows from `r paste(x)` where there are matching values in `r paste(x)`, and all columns from `r paste(x)` and `r paste(y)`.
```{r}
library(dplyr)
inner_join(itinerary, lodging, by = "city")
```

```{r}
inner_join(itinerary[2:5,], lodging[1:3,], by = "city") 
```

Returns all rows from `r paste(x)`, and all columns from `r paste(x)` and `r paste(y)`. Rows in `r paste(x)` with no match in `r paste(y)` will have NA values in the new columns.
```{r}
left_join(itinerary[2:5,], lodging, by = "city") 
```

```{r}
left_join(itinerary, lodging[2:5,], by = "city") 
```

Returns all rows from `r paste(y)`, and all columns from `r paste(x)` and `r paste(y)`. Rows in `r paste(y)` with no match in `r paste(x)` will have NA values in the new columns. 
```{r}
right_join(itinerary, lodging[2:5,], by = "city") 
```

Returns all rows from `r paste(x)` where there are matching values in `r paste(y)`, keeping just columns from `r paste(x)`.
```{r}
semi_join(itinerary, lodging[2:4,], by = "city") # returns only cols from itinerary 
```

Return all rows from `r paste(x)` where there are not matching values in `r paste(y)`, keeping just columns from `r paste(x)`.
```{r}
anti_join(itinerary[1:3,], lodging[3:4,], by = "city") # returns cols fron itinerary where there are no matching values in lodging 
```

Return all rows and all columns from both `r paste(x)` and `r paste(y)`. Where there are not matching values, returns NA for the one missing.
```{r}
full_join(itinerary[1:3,], lodging[3:6,], by = "city") 
```
