---

---

# Find character strings 

Character strings can be matched and manipulated in base R by using `regular expressions` in functions `grep`, `grepl`, `sub`, `gsub`, `regexpr` + `regmatches` and some others. 
Also tidyverse package 'stringr' contains analogous verbs with more consistent syntax.
A [regular expression](https://stat.ethz.ch/R-manual/R-devel/library/base/html/regex.html) is a pattern that describes a set of strings.

<!-- + The fundamental building blocks are the regular expressions that match a single character.  -->
+ Most characters, including all letters and digits, are regular expressions that match themselves. Whereas, e.g. `.` matches any single character.
+ You can refer also to a character class, which is a list of characters enclosed between `[` and `]`, e.g. `[[:alnum:]]` is same as `[A-z0-9]`.
+ Most common character classes: `[:alnum:]`, alphanumeric (`[:alpha:]` and `[:digit:]`); `[:alpha:]`, alphabetic (`[:upper:]` and `[:lower:]`); `[:punct:]`, ! " # $ % & ' ( ) * + , - . / : ; < = > ? @ [ \ ] ^ _ ` ` { | } ~.; `[:blank:]`, space and tab; etc. [Regular Expressions as used in R](https://stat.ethz.ch/R-manual/R-devel/library/base/html/regex.html)
+ The metacharacters in regular expressions are `. \ | ( ) [ { ^ $ * + ?`, whether these have a special meaning depends on the context. 
+ When matching any metacharacter as a regular character, precede it with a double backslash `\\`.
+ Repetition quantifiers put after regex specify how many times regex is matched: `?`, optional, at most once; `*`, zero or more times; `+`, one or more times; `{n}`, n times; `{n,}`, n or more times; `{n,m}`, n to m times. 
+ The caret `^` and the dollar sign `$` are metacharacters that respectively match the empty string at the beginning and end of a line.


To get the length of a text string (i.e. the number of characters in the string):
```{r}
x <- "reformation"
nchar(x)
```

`regexpr` returns position of 1st match and its length in a string:
```{r}
regexpr("o", x)
```
We can see that "o" is the 4th character in a string, with length 1 chracter (useBytes=T).

Lets try vector of strings. We want to return matching strings after underscore:
```{r}
y <- c("AreaShape_Area", "AreaShape_Perimeter", "AreaShape_MajorAxisLength", "Intensity_MADIntensity")
m <- regexpr("[[:alnum:]]*$", y) # to return start and length of match
m # to have have a look
```

What the heck we do with this list? With the help of `regmatches` command, you can return matching strings from `regexpr` call:
```{r}
regmatches(y, m) # to return matching strings
```

> Tidyverse's `stringr` library resolves the problem of difficult to work with putputs and produces outputs than can easily be used as inputs.

`stringr` library verbs to fulfill tasks by `regexpr` and `regmatches` are `str_locate` and `str_extract`:
```{r}
library(stringr)
str_locate(y, "[[:alnum:]]*$") # if we want only start and end positions
```
Note that, `str_locate` returns more conveniently a matrix (instead of a list as in regexpr).

```{r}
str_extract(y, "[[:alnum:]]*$") # if we want actual strings
```
Seems nice and tidy.

To return multiple matches from the string, not just the first one, use `gregexpr`, which returns start position and length of every match:
```{r}
regmatches(y, gregexpr("Area", y)) # to return start position and length of every match
```

Please try to return positions of all upper case letters from y:
```{r}
regmatches(y, gregexpr("[[:upper:]]", y))
```

Tidyverse alternatives to `gregexpr` is `str_locate_all` and `str_extract_all`:
```{r}
str_locate_all(y, "e") # to locate all start positions and lengths of letter "e"
```

```{r}
str_extract_all(y, "[[:upper:]]") # to extract all upper case letters
```

`grep(pattern, x)` finds a pattern in x. Pattern -- character string containing a regular expressions (or exact character string for fixed = TRUE).

What really happens when we use `grep` or `grepl` on a vector x.
By using `grep()` with default arguments gets us vector of the indices of the matching items in x:
```{r}
grep("Area", y)
```

Using `grep(value = TRUE)` we get character vector containing the selected elements of x:
```{r}
grep("Area", y, value = TRUE)
```

Tidy way to run `grep` is to use `str_subset()` from `stringr` library:
```{r}
str_subset(y, "Area") # note that strings and pattern have changed positions
```

However, by using `grepl` we get logical vector showing match for each element:
```{r}
grepl("Area", y)
```

Neat thing with the `grepl` is that it's versatile: you can use its result for subsetting or to count matches (TRUE = 1 and FALSE = 0):
```{r}
grepl("Area", y) %>% sum
```

Respectively, `stringr` library verb is `str_detect()`:
```{r}
str_detect(y, "Area")
```


We can use `grepl` in `dplyr::filter` to subset data frame. For example we want to look at the participants with first name 'Jaan' in 2013 Tartu Ski Marathon finish protocol:
```{r}
tm_2013 %>% filter(grepl(", Jaan", Nimi)) %>% tbl_df # Nimi is name and is given as 'lastname, firstname'.
```
Ups, we got 'Jaanus'-es too!

Let's add end of a line anchor `$` to get only 'Jaan's and lets use `str_detect` verb:
```{r}
tm_2013 %>% filter(str_detect(Nimi, ", Jaan$")) %>% tbl_df # Nimi is name and is given as 'lastname, firstname'. We use tbl_df to print only head of the table.
```
Good!

To substitute parts of a string, base R uses `sub` and `gsub` to perform replacement of the first and all matches respectively. `strinr` versions of these verbs are `str_replace` and `str_replace_all`, respectively:
```{r}
str_replace(y, "Area", "XXX") # to replace first occurence of Area in each string with XXX
```

```{r}
str_replace_all(y, "Area", "XXX") # replaces all Area-s with XXX
```
