---

---

<!-- # Using Dates and Times in R -->
<!-- % Bonnie Dixon -->
<!-- % 14-02-10 15:09:57 -->

<!-- *Today at the [Davis R Users' -->
<!-- Group](http://www.noamross.net/davis-r-users-group.html), [Bonnie -->
<!-- Dixon](http://ffhi.ucdavis.edu/people/directory/bmdixon) gave a tutorial on the -->
<!-- various ways to handle dates and times in R. Bonnie provided this great script -->
<!-- which walks through essential classes, functions, and packages. Here it is piped through -->
<!-- `knitr::spin`. The original R script can be found as a gist -->
<!-- [here](https://gist.github.com/noamross/8928124).* -->

# Date and time classes
Three date and time classes are built-in in R, `Date`, `POSIXct`, and `POSIXlt`.

## Date
If you have **only dates** in your data (but no times).

Create a date:
```{r}
bs <- as.Date("2016-08-29")
class(bs)
bs
```
If you don't specify format, `as.Date` will try "%Y-%m-%d" then "%Y/%m/%d". Other formats must be specified:

```{r}
es <- as.Date("01/29/2017", format = "%m/%d/%Y")
es  
```

```{r}
ch <- as.Date("December 19, 2016", format = "%B %d, %Y")
ch
```

The list of format symbols can be accessed via `strptime`:
```{r, eval=FALSE}
?strptime     
```

Get the current date:
```{r}
Sys.Date()
```

**Calculations with dates**. Find the difference between dates:

```{r}
es - bs
```


```{r}
difftime(ch, bs, units = "weeks")
```

Add or subtract days:
```{r}
bs + 40
```


```{r}
ch - 10
```

Create a vector of dates and find the intervals between them:

```{r}
road.fatalities <- as.Date(c("01.01.2016", "05.01.2016", "11.01.2016","18.01.2016","02.02.2016","08.02.2016","19.02.2016","25.02.2016", "02.03.2016", "28.03.2016", "29.03.2016", "21.03.2016", "04.04.2016", "12.04.2016", "13.04.2016","09.05.2016","13.05.2016","16.05.2016", "01.08.2016","22.08.2016"), format = "%d.%m.%Y")
road.fatalities <- sort(road.fatalities) # just in case, lets sort it
road.fatalities
```

```{r}
diff(road.fatalities)
```

Create a sequence of dates:
```{r}
six.weeks <- seq(Sys.Date(), length = 6, by = "week") 
six.weeks
```

```{r}
six.weeks <- seq(Sys.Date(), length = 6, by = 7)
six.weeks
```

```{r}
six.weeks <- seq(Sys.Date(), length = 3, by = "2 weeks")
six.weeks
```

See the internal integer representation:
```{r}
unclass(bs)
```

Note, that internal integer representation of Date class is number of days from `January 1, 1970`:
```{r}
bs - as.Date("1970-01-01") # 
```

## POSIXct
If you have **times in your data**, this is usually the best class to use.

Create some POSIXct objects:
```{r}
mft1 <- as.POSIXct("2014-03-11	17:14:15")
mft1
```

Specify format:
```{r}
mft2 <- as.POSIXct("22-Oct-2013 16:30:55", format = "%d-%b-%Y %H:%M:%S")
mft2
```

Specify the time zone:
```{r}
rstudio.webinar <- as.POSIXct("November 9, 2016 11am", 
                  format = "%B %d, %Y %I%p", 
                  tz = "EST")
rstudio.webinar
```

**Some calculations with times**
Compare times:

```{r}
mft1 > mft2
```

Add or subtract seconds:
```{r}
mft1 + 30
```

```{r}
mft2 - 30
```

Find the difference between times:
```{r}
mft1 - mft2
```

Automatically adjusts for daylight savings time. Last sunday in march, clocks are adjusted forward one hour:
```{r}
as.POSIXct("2016-03-27 3:00:00") - as.POSIXct("2016-03-26 23:55:00") # note that time difference is only 2.08 hours
```

Get the current time (in POSIXct by default):
```{r}
Sys.time()
```

See the internal integer representation in seconds:
```{r}
unclass(mft1)
```

```{r}
difftime(mft1, as.POSIXct("1970-01-01 00:00:00", tz = "UTC"), units = "secs")
```

## POSIXlt
This class enables easy extraction of specific componants of a time. 

+ `ct` -- calendar time 
+ `lt` -- local time. `lt` also helps one remember that POXIXlt objects are *lists*.)

Create a time:
```{r}
mft1.lt <- as.POSIXlt("2014-03-11	17:14:15")
mft1.lt
```


```{r}
unclass(mft1.lt)
```

```{r}
unlist(mft1.lt)
```

Extract componants of a time object:
```{r}
mft1.lt$sec
```

```{r}
mft1.lt$wday
```

Truncate or round off the time:
```{r}
trunc(mft1.lt, "days")
```

```{r}
trunc(mft1.lt, "mins")
```

**Summary of date and time classes**

+ When you just have dates, use `Date`.
+ When you have times, `POSIXct` is usually the best,
+ `POSIXlt` enables easy extraction of specific components
+ and there is also `chron` package when you don't need to deal with timezones and daylight savings time.

## Manipulating times and dates

### lubridate
This package is a wrapper for POSIXct with more intuitive syntax.


```{r}
library(lubridate)
```

Create a time. Note that lubridate uses UTC time zones as default:
```{r}
mft1.lub <- ymd_hms("2014-03-11	17:14:15")
mft1.lub
```
We have to set timezone explicitly:
```{r}
mft2.lub <- dmy_hm("22-Oct-2013 16:30", tz = "EET") 
mft2.lub
```

Timezone can be easily changed using `force_tz`:
```{r}
mft1.lub <- force_tz(mft1.lub, "EET")
mft1.lub
```

```{r}
rstudio.webinar <- mdy_h("November 9, 2016 11am")
rstudio.webinar
```

```{r}
today <- ymd(Sys.Date())
today
```


Some manipulations:
Extract or reassign componants:
```{r}
year(mft1.lub)
```

```{r}
week(mft1.lub)
```

```{r}
wday(mft1.lub, label = TRUE)
```

```{r}
hour(mft1.lub)
```

```{r}
tz(mft1.lub)
```

```{r}
second(mft1.lub) <- 7
mft1.lub
```

Converting to decimal hours can facilitate some types of calculations:
```{r}
mft1.dechr <- hour(mft1.lub) + minute(mft1.lub)/60 + second(mft1.lub)/3600
mft1.dechr
```

+ Lubridate distinguishes between four types of objects: instant, interval, duration, and period.
+ An instant is a specific moment in time.
+ Interval, duration, and period are all ways of recording time spans.

Dates and times parsed in lubridate are instants:
```{r}
is.instant(mft1.lub)
```

Round an instant:
```{r}
round_date(mft1.lub, "minute")
```

```{r}
round_date(mft1.lub, "day")
```

Get the current time or date as an instant (`Sys.time()`):
```{r}
now()
```

Shortcut for `Sys.Date()`
```{r}
today()
```
Some calculations with instants. Note that the units are seconds:
```{r}
mft1.lub - mft2.lub
```

```{r}
mft1.lub > mft2.lub
```
Add seconds:
```{r}
mft1.lub + 30
```

An interval is the span of time that occurs between two specified instants.

```{r}
time.to.xmas <- interval(now(), dmy("24-12-2016"))
time.to.xmas
```

Check whether a certain instant occured with a specified interval:
```{r}
rstudio.webinar %within% time.to.xmas
```

```{r}
mft1.lub %within% time.to.xmas
```

Determine whether two intervals overlap:
```{r}
daylight <- as.interval(ymd_hm("2016-03-26 23:55"), ymd_hm("2016-03-27 3:00"))
daylight
```

```{r}
inbed <- as.interval(ymd_hm("2016-03-26 22:30"), ymd_hm("2016-03-27 7:30"))
```


```{r}
int_overlaps(inbed, daylight)
```

A **duration is a time span not anchored to specific start and end times**. It has an exact, fixed length, and is stored internally in seconds.
Lets go to the wild side and create some durations:
```{r}
six.minutes <- dminutes(6)
six.minutes
```

```{r}
five.days <- ddays(5)
five.days
```

And now, after some practice with days and minutes, lets create whole year...
```{r}
one.year <- dyears(1)
one.year
```

```{r}
as.duration(inbed)
```

Calculations with durations:
```{r}
year(mft1.lub)
mft1.lub + one.year
```

```{r}
five.days + dhours(12)
```

```{r}
six.minutes/as.duration(inbed)
```

A **period is a time span not anchored to specific start and end times and measured in units larger than seconds with inexact lengths**. 
Create some periods:
```{r}
three.weeks <- weeks(3)
three.weeks
```

```{r}
four.hours <- hours(4)
four.hours
```

Calculations with periods:
```{r}
mft2.lub
mft2.lub + three.weeks
```

```{r}
sabbatical <- months(6) + days(12)
sabbatical
```

```{r}
three.weeks/sabbatical
```

## Calculating mean clock times
Say we have a vector of clock times in decimal hours and **we want to calculate the mean clock time**.

```{r}
bed.times <- c(23.9, 0.5, 22.7, 0.1, 23.3, 1.2, 23.6)
bed.times
```

```{r}
mean(bed.times)  # doesn't work, gives arithmetic mean
```

The **clock has a circular scale**, which ends where it begins, so we need to use circular statistics.
Get the package `psych`.
```{r}
library(psych)
circadian.mean(bed.times)
```

## Using times and dates in a data frame
Here is a data frame with a week of hypothetical times of going to bed and getting up for one person, and the total amount of time sleep time obtained each night according to a sleep monitoring device.


```{r}
sleep <- 
  data.frame(
    bed.time = ymd_hms("2013-09-01 23:05:24", "2013-09-02 22:51:09",
                       "2013-09-04 00:09:16", "2013-09-04 23:43:31",
                       "2013-09-06 00:17:41", "2013-09-06 22:42:27",
                       "2013-09-08 00:22:27"),
    rise.time = ymd_hms("2013-09-02 08:03:29", "2013-09-03 07:34:21",
                        "2013-09-04 07:45:06", "2013-09-05 07:07:17",
                        "2013-09-06 08:17:13", "2013-09-07 06:52:11",
                        "2013-09-08 07:15:19"),
    sleep.time = dhours(c(6.74, 7.92, 7.01, 6.23, 6.34, 7.42, 6.45))
              ); sleep
```

We want to calculate sleep efficiency, 
 the percent of time in bed spent asleep.
```{r}
sleep$efficiency <- 
  round(sleep$sleep.time/(sleep$rise.time - sleep$bed.time)*100, 1)
sleep
```

Now let's calculate the mean of each column:
```{r, eval=FALSE}
colMeans(sleep)  # doesn't work
```

```{r}
circadian.mean(hour(sleep$bed.time) + 
               minute(sleep$bed.time)/60 + 
               second(sleep$bed.time)/3600)
```

```{r}
circadian.mean(hour(sleep$rise.time) + 
               minute(sleep$rise.time)/60 + 
               second(sleep$rise.time)/3600)
```


```{r}
mean(sleep$sleep.time)/3600
```

```{r}
mean(sleep$efficiency)
```

<!-- We can also plot sleep duration and efficiency across the week: -->
<!-- ```{r} -->
<!-- par(mar = c(5, 4, 4, 4)) -->
<!-- plot(round_date(sleep$rise.time, "day"), sleep$efficiency,  -->
<!--      type = "o", col = "blue", xlab = "Morning", ylab = NA) -->
<!-- par(new = TRUE) -->
<!-- plot(round_date(sleep$rise.time, "day"), sleep$sleep.time/3600,  -->
<!--      type = "o", col = "red", axes = FALSE, ylab = NA, xlab = NA) -->
<!-- axis(side = 4) -->
<!-- mtext(side = 4, line = 2.5, col = "red", "Sleep duration") -->
<!-- mtext(side = 2, line = 2.5, col = "blue", "Sleep efficiency") -->
<!-- ``` -->

<!-- More resources on times and dates -->
<!-- Date and time tutorials for R: -->

<!-- + [http://www.stat.berkeley.edu/classes/s133/dates.html](http://www.stat.berkeley.edu/classes/s133/dates.html) -->
<!-- + [http://science.nature.nps.gov/im/datamgmt/statistics/r/fundamentals/dates.cfm](http://science.nature.nps.gov/im/datamgmt/statistics/r/fundamentals/dates.cfm) -->
<!-- + [http://en.wikibooks.org/wiki/R_Programming/Times_and_Dates](http://en.wikibooks.org/wiki/R_Programming/Times_and_Dates) -->

<!-- [lubridate](http://www.jstatsoft.org/v40/i03/paper) -->

<!-- Time zone and daylight saving time info: -->

<!-- + [http://www.timeanddate.com/](http://www.timeanddate.com/) -->
<!-- + [http://en.wikipedia.org/wiki/List_of_tz_database_time_zones](http://en.wikipedia.org/wiki/List_of_tz_database_time_zones) -->
<!-- + [http://www.twinsun.com/tz/tz-link.htm](http://www.twinsun.com/tz/tz-link.htm) -->
<!-- + Also see the R help file at `?Sys.timezone` -->

