
# Functions

A function is a piece of code to carry out a specified task, e.g. calculate mean, split data frame. 
It may accept arguments (or not) and it may return zero to more values or objects.

General form of R function is following:

<pre><code>
function(argslist) expr
</code></pre>

Where, `argslist` is either empty or one to more `name` or `name=expression` terms.
When calling a function you can specify arguments by position, by complete name, or by partial name. 

Example function with no arguments, doing only one thing:
```{r}
dedication <- function() {
  cat("And Pierce Brosnan how dare you prescribe\n
      Sad grief and bed wet pills")
}
```

Above defined function prints some cryptic lyrics (Credit: ME Smith)
```{r}
dedication()
```

But we can tune above function to insert your computer's username into that line instead of that actor:
```{r}
not_medication <- function() cat(paste("And", system("whoami",T), "how dare you prescribe\n
Sad grief and bed wet pills"))
```

```{r}
not_medication()
```

+ Function names must be concise, [like above](http://thefall.org/discography/data/album31.html)
+ Note that, when writing a multiline function, braces `{}` must be used to enclose the body of the function.

Onliners like this does not need braces:

```{r}
plusone <- function(x) x + 1
plusone(1)
```

## Default arguments

Some arguments have default values specified, as shown below. Arguments without a default must have a value supplied for the function to run. You do not need to provide a value for those arguments with a default, as the function will use the default value:
```{r}
sum_stat <- function(x, f = "mean"){
  do.call(f, list(x))
}

v <- rnorm(10)
sum_stat(v) # by default this function calculates mean
```

But we can also calculate median, if we change f argument:
```{r}
sum_stat(v, f = "median") # 
```


## Ellipsis

The addition of `...`, or **ellipsis**, in the function definition allows other arguments to be passed into the function, and passed onto to another function:
```{r}
# Let's redifine sum_stat function with ellipsis to pass additional arguments to summary function
sum_stat2 <- function(x, f = "mean", ...){
  do.call(f, list(x, ...)) # do.call constructs and executes a function call from a function and a list of arguments to be passed to it
}

v2 <- c(NA, v, NA, NA)
sum_stat2(v2) # NA-s start to give trouble, as mean fun has na.rm = FALSE argument

sum_stat2(v2, na.rm = TRUE) # here we pass na.rm = TRUE to our function and voila!
```



## Function environment
By default, function returns result from last expression:
```{r}
logit <- function(p) {
 odds <- p/(1-p)
 log(odds)
}

logit(0.5)
```

Importantly, objects that are created within the function are local to the environment of the function. If you direct your last expression into an object inside function, you need to explicitly `return()` the value of that object:
```{r}
# look what happens
logit2 <- function(p) {
 odds <- p/(1-p)
 log_odds <- log(odds) # last expression's value is assigned to log_odds object inside function
}
logit2(0.5) # nothing is returned

## use of return()
logit3 <- function(p) {
 odds <- p/(1-p)
 log_odds <- log(odds) # last expression's value is assigned to log_odds object inside function
 return(log_odds) # return object log_odds value
 }
logit3(0.5) 

## use of return()
logit4 <- function(p) {
 odds <- p/(1-p)
 log_odds <- log(odds) # last expression's value is assigned to log_odds object inside function
 return(list(odds = odds, log_odds = log_odds)) # return two objects from function environment as named list
}

logit4(0.5) 
```

## Anonymous function inside for loop
`lapply()`/`sapply()` takes a function, applies it to each element in a list/vector, and returns the results in the form of a list. Its easy to use when you want to calculate eg. mean of each list element. But what if you need to do something more specific with your data:
```{r}
let <- c("a","z","x","u","b") # we have a list of letters
sapply(let, function(x) grep(x, letters)) # we need to know the position of each letter in alphabet
```

## Functional sequence
`magrittr` allows to use `%>%` piping operator not only to produce values but also [to produce functions](https://blog.rstudio.org/2014/12/01/magrittr-1-5/). This kind of functional sequences denote input with `.` dot:

```{r}
library(magrittr)
mae <- . %>% abs %>% mean(na.rm = TRUE)
mae(rnorm(10))
```

This is similar to following function:
```{r, eval=FALSE}
mae <- function(x) {
  mean(abs(x), na.rm = TRUE)
}
```

Quite common use case of these functional sequences is inside lapply/sapply:

```{r, eval=FALSE}
library(dplyr)
library(readr)
library(reshape2)
data <- paths-to-csv-files %>% lapply(. %>% read_csv %>% melt)
```

It's cool that you can subset these functional sequences and use only parts of them when necessary:
```{r}
mae[2](rnorm(5))
mae[1](rnorm(10))
```


### Another example

Let's create function to calculate __standardised effect size, Cohen's d__. First let's simulate some toy data of blood pressure readings from patients with high systolic bp, treated either with drug or placebo: 
```{r, fig.width=3}
library(ggplot2)
probs_h <- pnorm(70:190, 150, 7) 
probs_n <- pnorm(70:190, 130, 7)
m <- data.frame(bp = c(sample(70:190, 56, replace = T, prob = probs_h), 
                sample(70:190, 60, replace = T, prob = probs_n)), 
                tr = c(rep("placebo", 56), rep("drug", 60)))
```

This is how our dummy dataset looks like:
```{r, fig.width=3}
ggplot(m, aes(tr, bp)) + geom_boxplot() + geom_jitter(width = 0.5)
```


[Cohen's d is defined](https://en.wikipedia.org/wiki/Effect_size) as the difference between two means divided by a standard deviation for the data: 

$$d = \frac{\bar{x}_1-\bar{x}_2}{s}$$

$s$ is pooled standard deviation:

$$s = \sqrt{\frac{(n_1-1){s_1}^2+(n_2-1){s_2}^2}{n_1+n_2-2}}$$ 

Using above equations, we define function:
```{r}
cohensd <- function(x1, x2){ # '{}' because our function is spread on multiple lines
  
  n1 <- length(x1) - 1
  n2 <- length(x2) - 1
  
  # mean difference
  md  <- abs(mean(x1, na.rm = T) - mean(x2, na.rm = T))        
  
  # common sd
  s <- n1 * var(x1, na.rm = T) + n2 * var(x2, na.rm = T)
  s <- s/(n1 + n2)
  s <- sqrt(s)                     
  
  # calculate Cohen's d
  md/s # last expression's result is returned
}
```

Cohen's d is interpreted that d values of 0.2, 0.5, and 0.8 represent small, medium, and large effect sizes respectively:
```{r}
bp <- split(m$bp, m$tr) # to split data frame into two vectors
cohensd(bp$placebo, bp$drug) # to input treatment groups as vectors
```
Because our function arguments are two vectors x1 and x2 and we had data frame in the long format, we had to split data frame into vectors. [Can we embed these two vectors into data frame in the long format (two cols - one col per vector)?] 

We can try to wrap this function into another function to use our data frame:
```{r}
cohens_d <- function(data, values, group){
  v <- split(data[[values]], data[[group]])
  cohensd(v[[1]], v[[2]])
}
save(cohens_d, file = "lib/cohens_d.R") # save your function into lib folder
```

```{r}
cd <- cohens_d(m, "bp", "tr")
cd
```

Now we can create function to insert word explaining the size of the effect into text:
```{r}
mansplain_cohensd <- function(x) {
  if(x<=0.2) return("small") 
  if(x>=0.2&x<=0.5) return("medium")
  if(x>0.5&x<Inf) return("large")
  if(x==Inf) return("cosmos")
}

cdv <- mansplain_cohensd(cd)
cdv
```

```
Seems that the standardised effect size for the drug treatment effect on blood pressure is **` r round(cd, 2)`**, which means it's **` r mansplain_cohensd(cd)`**.
```
Whitch produces following output:

Seems that the standardised effect size for the drug treatment effect on blood pressure is **`r round(cd, 2)`**, which means it's **`r mansplain_cohensd(cd)`**.

-----------------------------------------------------------------------------------
+ Try to keep your functions in separate sub directory `lib` in your project's dir 
+ load functions using `load(lib/cohens_d.R)`

Good function is:

+ short
+ performs one task
+ has intuitive short name
